An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-02-03 rule-ltl.tex
fixing problems

@@

@@

We now extend FLTL into EFLTL in a way that was motivated by the the propositional extension from LTL (PLTL) to ELTL (PELTL). We thus allow the following formula:
\begin{equation}
\label{EFLTL}
\psi \mathit{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \end{equation}
\begin{enumerate}
\item $\psi$, the {\em statement}, is a FLTL formula with
no free variables, 
\item $\varphi_j$ are PFLTL formula with a single
free variable $x_j$,
\item $r_j$ is an auxiliary temporal relations with $2$ parameters; the first parameter is of the same 
type as $x$ and the second one is, as usual in this paper,
a natural number that is omitted in the temporal formulas. 
An auxiliary relations $r_j$ can appear within $\psi$. It can also appears in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, only within the
scope\footnote{This restricted has the same motivation as
the propositional version ELTL (and EPLTL): to avoid making the
formula $\false$ just by means of the definition of the auxiliary relations.}
of a previous-time operator $\ominus$. 
\end{enumerate}

Again, it
is simple to extend the syntax and semantics to include
relations $r_j$ with different number of arguments, and, correspondingly, formulas $\varphi_j$ with corresponding number of free
variables\footnote{We chose not to do that here (and in the definition of the semantics of FLTL above), because it
requires multi-indexing notation that is trivial and adds no further insight. Our implementation indeed uses
this more general capabilities.}.
The logic EPFLTL is obtain by restricting the temporal modalities of EFLTL to the past ones:
$\since$ and $\ominus$, and those obtained from them.

@@

We define\footnote{This is done for lack of space. Formal semantics 
can be given by constructing a set of temporal relation extended
with the auxiliary ones
inductively, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL specification by using the following equivalent
QFLTL formula:
\begin{equation} \label{equiv}
\Exists r_1  \, \ldots \, \Exists r_n \,  ( \psi \wedge \Box \bigwedge_{j \in \{ 1, \ldots , n\}} (r_j ( x_j )  \leftrightarrow 
\varphi_i ( x_j) )
\end{equation}

The analogous of Lemma~\ref{simplecase} shows that EPFLTL does not
need the past temporal operators besides $\ominus$.
\begin{lemma}
The full expressive power of EPFLTL can be obtained by allowing only 
specifications of the form
\[ \Box \psi \mathrm{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} \]
where $\psi$ consists of a fully quantified single occurrence of some
auxiliary relation $r_j$.
\end{lemma}
\noindent {\bf Proof.} Basically, the proof is the same as Lemma~\ref{simplecase}, based on the axiomatization of the
past temporal operators. \qed

%\klaus{Something seems wrong above.}

@@

We showed in Theorem~\ref{theo1} that although ELTL
can be translated into QLTL, both have the same
expressive power. We study now the relationship between
EFLTL and QFLTL.

@@

Consider again the EPFLTL form
of specification, described in~(\ref{EFLTL}), restricted to the past operators. As syntax and semantics of various versions
of LTL are given in this paper, we repeat now the concrete definitions for EPFLTL. The syntax is as follows:

\begin{center}
$\varphi ::= \true  \; | \;
    p ( a ) \; | \;
    p ( x ) \; | \;
    ( \varphi \wedge \varphi ) \;  |   \;
   \neg \varphi \; | \;
    \ominus \varphi \; | \;
    ( \varphi  \; \since  \; \varphi ) \; | \;
    \exists x \; \varphi$
\end{center}
\[ 
\psi \mathrm{\ where\ } r_j ( x_j) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \}
\]

@@

@@

\begin{itemize}
\item $I [ \varphi , {\cal R} , 0 ] = \emptyset$.
\item $I [ \true , {\cal R} , i ] = \{ \epsilon \}$.
\item $I [ p ( a ) , {\cal R} , i ] =$ if $p ( a ) \in \sigma [ i ]$ then
$\{ \epsilon \}$ else $\emptyset$.
\item $I [ p ( x ) , {\cal R} , i ] = \{ [ x \mapsto a ] | p ( a ) \in
\sigma [ i ] \}$.
%\item $I [ \true , \sigma , i ] = D^n$.
%\item $I [ ( \varphi \vee \psi ) , \sigma , i ] = 
%I [ \varphi , \sigma , i ] \cup I [ \psi , \sigma , i ]$.
\item $I [ ( \varphi \wedge \psi ) , \sigma , i ] = 
I [ \varphi , {\cal R} , i ] \;  \bigcap \; I [ \psi , \sigma , i ]$.
\item $I [ \neg \varphi , {\cal R} , i ] = 
A_{\vars ( \varphi )} \; \setminus \; I [ \varphi , {\cal R} , i ]$.
\item $I [ ( \varphi \; \since \; \psi ) , {\cal R} , i ] = 
I [ \psi , {\cal R}, i ] \; \bigcup \;
( I [ \varphi , {\cal R} , i ] \bigcap 
I [ (\varphi \since \psi ) , {\cal R} , i - 1 ] )$.
\item $I [ \ominus \varphi , {\cal R} , i ] = I [ \varphi , {\cal R} , i-1 ]$.
\item $I [ \exists x \; \varphi , \sigma , i ] = 
\proj ( I [ \varphi , {\cal R} , i ] , \{ x \} )$.
\item $I [ \psi \mathrm{\ where\ } r_j ( x_j) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} , {\cal R} , i ] = 
 I [ \psi , {\cal Q} ] $ where ${\cal Q} = {\cal R} \cup
\{ r_j | r_j [ i ] =  I [ 
\varphi_j , {\cal Q} , i ] \}$
\end{itemize}
In the last item, the extended model ${\cal Q}$ appears both on the left and the right of the equality sign. It is well defined because of Lemma~\ref{samesame}.

We describe an algorithm for monitoring 
FLTL properties, first presented in~\cite{HPU} and implemented
in the tool~\dejavu. 
Instead of storing the data values occurring in events, 
we enumerate these data values as soon as we see them and use Boolean
encodings of this enumeration.
For example, if the runtime verifier sees the input 
events 
$\mathit{open}(\text{``a''})$, 
$\mathit{open}(\text{``b''})$, 
$\mathit{open}(\text{``c''})$, 
it will encode them as
$000$, $001$ and $010$ (say, we use 3 bits $b_0$, $b_1$ and $b_2$
to represent each enumeration, with $b_2$ being the most significant bit).

@@

Since we want to be able to deal with infinite domains
(where only a finite number of elements may appear in a given observed prefix) and maintain the ability to perform
complementation, unused enumerations represent the
values that have not been seen and their relations
to all other values. The algorithm maintains that these unseen values are correctly represented, as shown in~\cite{HPU}. In fact, it is sufficient to have just one enumeration representing these values per each variable of the LTL formula. However, it is important to guarantee that at least one such enumeration
exists. We guarantee that by preserving the enumeration $11\ldots11$, i.e., the highest Boolean representation.
We present here only the basic algorithm. For versions that
allow extending the number of bits used for enumerations and garbage collection of enumerations, consult~\cite{HP}.

A Boolean representation of the {\em set} of values 
$\{\text{``a''},\text{``c''}\} $ would be equivalent to a Boolean function $(\neg b_0 \wedge \neg b_2)$ that returns 1 for $000$ and $010$ (the value
of $b_1$ can be arbitrary).

    Our preferred c of sets of values and
assignments is as an Ordered Binary Decision
Diagram (OBDD, although we write simply BDD)~\cite{Bryant}.
A BDD is essentially a compact representation 
of a Boolean tree, where compaction glues together isomorphic subtrees. Each non-leaf node is labeled with one of the
Boolean variables $b_0,\ldots,b_{k-1}$. A non-leaf node $b_i$ 
is the source of two 
arrows leading to other nodes. 
%A dotted-line arrow represents that $b_i$ has the Boolean value $0$, while a thick-line 
%arrow represents that it has the value $1$. 
The nodes in the DAG have the
same order along all paths from the root. However, some of the nodes may be
absent along some paths, when the result of the Boolean function does not 
depend on the value of the corresponding Boolean variable. Each path leads 
to a leaf node that is marked by either a $0$ or a $1$, representing the 
Boolean value returned by the function for the Boolean values on the path.

@@

>>>>

\klaus{I think that the structure of the paper should be
crystal clear so that the reader does not get lost. A lost reader leads to rejection. If a reader understand the structure, but perhaps not the details, the reader may still be happy.}

@@

\klaus{Below, has the notation ${\cal R}' |_{\{ q \} }$ been defined?}

@@

We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We thus allow the following formula:

\begin{equation}
\label{EFLTL}
\psi \mathit{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n 
\end{equation}

\begin{enumerate}
\item $\psi$, the {\em statement}, is a FLTL formula with
no free variables, 
\item $\varphi_j$ are PFLTL formulas with a single
free variable $x_j$,
\item $r_j$ is an auxiliary temporal relation with $2$ parameters; the first parameter is of the same 
type as $x_j$ and the second one is, as usual,
a natural number that is omitted in the temporal formulas. 
An auxiliary relation $r_j$ can appear within $\psi$. It can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
scope\footnote{This restriction has the same motivation as in
the propositional version ELTL (and EPLTL): to avoid making the
formula $\false$ just by means of the definition of the auxiliary relations.}
of a previous-time operator $\ominus$. 
\end{enumerate}

Again, it
is simple to extend the syntax and semantics to include
relations $r_j$ with different number of arguments, and, correspondingly, formulas $\varphi_j$ with corresponding number of free
variables\footnote{We chose not to do that here (and in the definition of the semantics of FLTL above), because it
requires multi-indexing notation that is trivial and adds no further insight. Our implementation indeed uses
this more general capability.}.
The logic EPFLTL is obtained by restricting the temporal modalities of EFLTL to the past ones:
$\since$ and $\ominus$, and those obtained from them.

@@

\noindent
We define\footnote{This is done for lack of space. Formal semantics 
can be given by constructing a set of temporal relations extended
with the auxiliary ones
inductively, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL 
specification (\ref{EFLTL}) 
by using the following equivalent
QFLTL formula:

\begin{equation} \label{equiv}
\Exists r_1  \, \ldots \, \Exists r_n \,  ( \psi \wedge \Box \bigwedge_{j \in \{ 1, \ldots , n\}} (r_j ( x_j )  \leftrightarrow 
\varphi_i ( x_j) )
\end{equation}

\noindent 
The analogous of Lemma~\ref{simplecase} shows that EPFLTL does not
need the past temporal operators besides $\ominus$.

\begin{lemma}
The full expressive power of EPFLTL can be obtained by allowing only 
specifications of the form
\[ \Box \psi \mathrm{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} \]
where $\psi$ consists of a fully quantified single occurrence of some
auxiliary relation $r_j$.
\end{lemma}

\klaus{Something seems wrong in the above formulation.
E.g.: Is the $\Box$ a restriction?, what is ``fully quantified single occurrence of some auxiliary relation $r_j$''?}

\noindent {\bf Proof.} Basically, the proof is the same as Lemma~\ref{simplecase}, based on the axiomatization of the
past temporal operators. \qed

@@

We showed in Theorem~\ref{theo1} that ELTL
and QLTL have the same
expressive power. We study now the relationship between
EFLTL and QFLTL.

@@

\klaus{In the text above there are some imprecisions, such as
(1) ``tuples of relations'' - relations are sets of tuples, we don't have tuples of relations. 
(2) ``${\cal R}[i]$ consists of some relations'': it is a set of
relations (I think).
(3) ``to have, per each state just some finite tuples'': tuples are finite. You mean: to contain a finite set of tuples?
(4) what is the story about a set of assignments also being a relation? Needed?
}

\klaus{Below: (\ref{EFLTL}) was not a EPFLTL form of specification.}

Consider again the EPFLTL form
of specification, described in~(\ref{EFLTL}), restricted to the past operators. We now present the definition of EPFLTL, the final main logic proposed in this paper. 
The syntax is as follows:

\[ 
\psi \mathrm{\ where\ } r_j ( x_j) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \}
\]
\begin{center}
$\varphi ::= \true  \; | \;
    p ( a ) \; | \;
    p ( x ) \; | \;
    ( \varphi \wedge \varphi ) \;  |   \;
   \neg \varphi \; | \;
    \ominus \varphi \; | \;
    ( \varphi  \; \since  \; \varphi ) \; | \;
    \exists x \; \varphi$
\end{center}

\klaus{Note that $\psi$ above is not defined as a syntactic category. It should really be $\varphi$. Likewise in other parts of the paper. Perhaps too big a change now? Also, there should be
a greek letter for a specification. E.g.:

\[ 
\pi ::= \varphi \mathrm{\ where\ } r_j ( x_j) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \}
\]
\[
\varphi ::= ...
\]
}

@@

\klaus{One may of course ask whether the following section on the 
set semantics is needed. Maybe it is needed, since we probably need a semantics. Something to consider.}

@@

\klaus{Above: assignments are introduced without any gentle introduction. Remember: you are not communicating with Moshe Vardi. You are communicating with John Fingelstein who had three glasses of wine at dinner.}

\klaus{Above: $I [ \varphi, {\cal R} , i]$ is introduced out of the blue without any explanation.}

\begin{itemize}
\item $I [ \varphi , {\cal R} , 0 ] = \emptyset$.
\item $I [ \true , {\cal R} , i ] = \{ \epsilon \}$.
\item $I [ p ( a ) , {\cal R} , i ] =$ if $p ( a ) \in \sigma [ i ]$ then
$\{ \epsilon \}$ else $\emptyset$.
\item $I [ p ( x ) , {\cal R} , i ] = \{ [ x \mapsto a ] | p ( a ) \in
\sigma [ i ] \}$.
%\item $I [ \true , \sigma , i ] = D^n$.
%\item $I [ ( \varphi \vee \psi ) , \sigma , i ] = 
%I [ \varphi , \sigma , i ] \cup I [ \psi , \sigma , i ]$.
\item $I [ ( \varphi \wedge \psi ) , \sigma , i ] = 
I [ \varphi , {\cal R} , i ] \;  \bigcap \; I [ \psi , \sigma , i ]$.
\item $I [ \neg \varphi , {\cal R} , i ] = 
A_{\vars ( \varphi )} \; \setminus \; I [ \varphi , {\cal R} , i ]$.
\item $I [ ( \varphi \; \since \; \psi ) , {\cal R} , i ] = 
I [ \psi , {\cal R}, i ] \; \bigcup \;
( I [ \varphi , {\cal R} , i ] \bigcap 
I [ (\varphi \since \psi ) , {\cal R} , i - 1 ] )$.
\item $I [ \ominus \varphi , {\cal R} , i ] = I [ \varphi , {\cal R} , i-1 ]$.
\item $I [ \exists x \; \varphi , \sigma , i ] = 
\proj ( I [ \varphi , {\cal R} , i ] , \{ x \} )$.
\item $I [ \psi \mathrm{\ where\ } r_j ( x_j) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} , {\cal R} , i ] = 
 I [ \psi , {\cal Q} ] $ where ${\cal Q} = {\cal R} \cup
\{ r_j | r_j [ i ] =  I [ 
\varphi_j , {\cal Q} , i ] \}$
\end{itemize}

\klaus{Above: as the semantics is presented, it looks as if 
the `$\psi\ where ...$' is just another subformula.}

\klaus{Above: hide is not defined.}

\klaus{Above: the semantics of $p(a)$ and $p(x)$ do not refer to 
${\cal R} $.}

In the last item, the extended model ${\cal Q}$ appears both on the left and the right of the equality sign. It is well defined because of Lemma~\ref{samesame}.

\subsubsection{Runtime Verification Algorithm for PFLTL}

We start by describing an algorithm for monitoring 
PFLTL properties, first presented in~\cite{HPU} and implemented
in the tool~\dejavu. 
Instead of storing the data values occurring in events, 
we enumerate these data values as soon as we see them and use Boolean
encodings of this enumeration, represented as BDDs.
For example, if the runtime verifier sees the input 
events 
$\mathit{open}(\text{``a''})$, 
$\mathit{open}(\text{``b''})$, 
$\mathit{open}(\text{``c''})$, 
it will encode them as
$000$, $001$ and $010$ (say, we use 3 bits $b_0$, $b_1$ and $b_2$
to represent each enumeration, with $b_2$ being the most significant bit).
%
A Boolean representation of the {\em set} of values 
$\{\text{``a''},\text{``b''}\} $ would be equivalent to a Boolean function $(\neg b_1 \wedge \neg b_2)$ that returns 1 for $000$ and $001$ (the value
of $b_0$ can be arbitrary).

@@

Since we want to be able to deal with infinite domains
(where only a finite number of elements may appear in a given observed prefix) and maintain the ability to perform
complementation, unused enumerations represent the
values that have not been seen and their relations
to all other values. The algorithm maintains that these unseen values are correctly represented, as shown in~\cite{HPU}. In fact, it is sufficient to have just one enumeration representing these values per each variable of the LTL formula. However, it is important to guarantee that at least one such enumeration
exists. We guarantee that by preserving the enumeration $11\ldots11$, i.e., the highest Boolean representation,
to represent values not yet seen.
We present here only the basic algorithm. For versions that
allow extending the number of bits used for enumerations and garbage collection of enumerations, consult~\cite{HP}.

Our preferred representation of a set of values 
(assignments) is as an Ordered Binary Decision
Diagram (OBDD, although we write simply BDD)~\cite{Bryant}.
A BDD is essentially a compact representation 
of a Boolean tree, where compaction glues together isomorphic subtrees. Each non-leaf node is labeled with one of the
Boolean variables $b_0,\ldots,b_{k-1}$ (where $k$ is the number of bits used to represent the values of a variable). A non-leaf node $b_i$ 
is the source of two 
arrows leading to other nodes, representing respectively that the node has the value 0 (false) or 1 (true).
%A dotted-line arrow represents that $b_i$ has the Boolean value $0$, while a thick-line 
%arrow represents that it has the value $1$. 
The nodes in the DAG have the
same order along all paths from the root. However, some of the nodes may be
absent along some paths, when the result of the Boolean function does not 
depend on the value of the corresponding Boolean variable. Each path leads 
to a leaf node that is marked by either a $0$ or a $1$, representing the 
Boolean value returned by the function for the Boolean values on the path.

@@

\subsubsection{Runtime Verification Algorithm for EPFLTL}
